% =========================================================================
% Pilot section v2 — replaces §5 Empirical Pilot and App. §B Pilot Operationalization
% from main.tex. Built against the v2 dataset (468 trials, 13 models, 36 items,
% 3 salience levels) rather than the v1 dataset (144 trials, 6 models, 24 items,
% 2 salience levels) the original section was written for.
%
% <<<...>>> slots are template placeholders to be filled from the final
% analyze.py --g-study output once the judge run completes. Each slot has
% the OLD value (from v1 pilot) in a comment for reference.
% =========================================================================

%% Drop-in replacement for the abstract's empirical sentence.
%% OLD: "A small-scale pilot against six open-weights models shows the partition
%%      is empirically realizable, prohibition salience is the largest single
%%      effect on violation rate (1.2% at high salience vs.\ 23.3% at low salience),
%%      and the HIUA-to-KBV ratio is approximately 2.75:1."

% NEW: A 13-model pilot spanning closed-API and self-hosted vLLM panels (468
% trials across a 36-item factorial) shows the $2{\times}4{\times}2$ partition is
% empirically realizable, prohibition salience produces a non-monotonic
% three-level effect on violation rate (<<<HIGH_VIOL>>>\% high vs.\ <<<MID_VIOL>>>\% mid
% vs.\ <<<LOW_VIOL>>>\% low, with mid the highest), and the recall probe remains
% effectively at ceiling, foreclosing direct observation of HIUA on this panel
% while sharpening the validity argument for the construct's measurement target.

% =========================================================================

\section{Empirical pilot}
\label{sec:pilot}

We pilot HIUA-Bench against a 13-model panel spanning two serving paths.
Six models were served through Groq's hosted inference endpoint (Llama 3.1 8B
Instant, Llama 3.3 70B Versatile, Llama 4 Scout 17B preview, Qwen3 32B
preview, OpenAI GPT-OSS 20B, and OpenAI GPT-OSS 120B). Seven additional
open-weights models were self-hosted via vLLM on Modal-managed GPUs to extend
the panel below and above Groq's catalog: Qwen 2.5 7B Instruct and Llama 3.1
8B Instruct on a single L40S; Gemma 2 27B Instruct and OpenAI GPT-OSS 20B on
a single A100-80GB; Llama 3.3 70B Instruct and Qwen 2.5 72B Instruct on 2
tensor-parallel H100s; and OpenAI GPT-OSS 120B on 4 tensor-parallel H100s.
The Llama 3.1 8B and 3.3 70B and the two GPT-OSS variants therefore appear in
the panel twice, once via each serving path, supporting a cross-provider
replication check on those four models. The full 36-item factorial covered 3
hallucination sub-constructs $\times$ 3 prohibition-salience levels $\times$ 4
domains, with one trial per (model, item) cell at temperature 0.7 with seed
0, yielding $13 \times 36 = 468$ trials. Of those, <<<N_ERRORS>>> failed at the
agent layer (the gpt-oss-20b tool-call rejection mode discussed below),
leaving <<<N_CLEAN>>> clean trials for the headline analysis.

For each trial, the agent received a system prompt containing the explicit
prohibition and a user message containing the hallucination trigger, then
produced a chain-of-thought followed by an action description. The recall
probe was issued in a fresh context. The action scorer applied a three-layer
cascade: a refusal-prefix heuristic on the parsed action line (extended to
recognize Harmony-format reasoning channels and polite-refusal openings used
by the GPT-OSS family); a regex match against forbidden and compliant action
patterns; and a fallback to a Llama 3.3 70B judge when the regex flagged both
forbidden and compliant matches. The recall scorer applied a two-layer
cascade: substring match against the expected recall token, and a fallback to
the same judge when the substring missed. Post-hoc, a three-judge ensemble
(Claude Sonnet 4.6, Llama 3.3 70B Versatile, and Llama 4 Scout 17B) labeled
each trial's hallucination type as one of $\{$none, authorization, state,
tool$\}$. Scoring-stack diagnostics and per-judge reliability are in
Appendix~\ref{app:pilot}.

\paragraph{Cell occupancy.}
Table~\ref{tab:pilot-cells} reports the headline $2 \times 2$ of
$\{$forbidden action?$\} \times \{$recalled rule?$\}$ pooled across the panel.
The recall probe was answered correctly on essentially every trial: the
panel knew the rule in nearly all cases. Violations occurred on
<<<VIOLATION_PCT>>>\% of clean trials, almost exclusively in the
\emph{violation-recalled} (knows-but-violates) cell.

\begin{table}[ht]
\centering
\caption{Pilot cell occupancy across $N=\langle$<<<N_CLEAN>>>$\rangle$ clean trials.}
\label{tab:pilot-cells}
\begin{tabular}{lrr}
\toprule
Cell & Count & Percent \\
\midrule
Compliant + recalled                       & <<<CR_N>>> & <<<CR_PCT>>>\% \\
Compliant + not recalled                   & <<<CNR_N>>> & <<<CNR_PCT>>>\% \\
Violation + recalled                       & <<<VR_N>>> & <<<VR_PCT>>>\% \\
Violation + not recalled (HIUA-candidate)  & <<<VNR_N>>> & <<<VNR_PCT>>>\% \\
\bottomrule
\end{tabular}
\end{table}

Of the <<<VR_N>>> violation-with-recall trials, the three-judge majority
hallucination label was \emph{authorization} on <<<MAJ_AUTH>>>,
\emph{state} on <<<MAJ_STATE>>>, \emph{tool} on <<<MAJ_TOOL>>>, and
\emph{none} on <<<MAJ_NONE>>>. Defining HIUA as the rate of forbidden action
conditional on recall and a hallucination-type majority label, and KBV as the
rate of forbidden action conditional on recall and a ``none'' majority label
in this restricted partition:
\[
\widehat{\mathrm{HIUA}} = \tfrac{<<<HIUA_NUM>>>}{<<<N_CLEAN>>>} = <<<HIUA_PCT>>>\%, \qquad
\widehat{\mathrm{KBV}} = \tfrac{<<<KBV_NUM>>>}{<<<N_CLEAN>>>} = <<<KBV_PCT>>>\%.
\]
The HIUA-to-KBV ratio is approximately <<<HIUA_KBV_RATIO>>>:1. Note that the
$\{$violation, not recalled$\}$ cell remains empty: HIUA in its
strongest construct definition---an action taken when the rule was not
recalled at probe time---is unobserved on this panel. The HIUA estimate
above therefore uses the looser definition adopted in the original DriftBench
partition, in which the recall criterion alone does not suffice and the
hallucination-type label is required to separate confabulated from
lucid mechanism. We return to this distinction in §\ref{sec:discussion}.

\paragraph{Per-model rates.}
Table~\ref{tab:pilot-per-model} reports per-model rates ($n=36$ each, except
on models where the agent layer errored, see App.~\ref{app:pilot}). The
panel spans a capability range from 7B parameters to 120B (with one
235B-active-parameter MoE in the self-hosted opus tier excluded from the
dev pilot). Within-family scaling is observable on Meta-Llama (3.1 8B vs.\
3.3 70B) and on OpenAI's open-weights family (GPT-OSS 20B vs.\ 120B).

\begin{table}[ht]
\centering
\caption{Per-model pilot rates. Cells in percent. HIUA-cand and KBV use the
three-judge majority hallucination label among the violation-with-recall
trials. Provider $\in \{$G (Groq), S (selfhost)$\}$.}
\label{tab:pilot-per-model}
\small
\begin{tabular}{llrrrrr}
\toprule
Model & Prov. & $n$ & Violation & Recall & HIUA-cand & KBV \\
\midrule
<<<MODEL_TABLE_ROWS>>>
\bottomrule
\end{tabular}
\end{table}

Cross-provider replication on the four models served both ways
(Llama 3.1 8B, Llama 3.3 70B, GPT-OSS 20B, GPT-OSS 120B) shows
<<<CROSS_PROVIDER_NOTE>>>.

\paragraph{Sub-construct and salience effects.}
The pilot's two largest single effects are reported in
Table~\ref{tab:pilot-subconstruct}. The salience effect is
\emph{non-monotonic}: violations are lowest at high salience (the
prohibition stated as an ABSOLUTE-RULE caps clause near the top of the
system role), highest at mid salience (the prohibition stated as one of
several operational sentences mid-paragraph), and intermediate at low
salience (the prohibition tucked into a tool docstring or parenthetical).
The mid-salience condition produces the highest violation rate not because
the rule is hardest to find---models can recall it at near-100\%
rates---but because its prose framing positions it as one of many
operational instructions to balance against the user request rather than as
the one absolute constraint to honor unconditionally.

\begin{table}[ht]
\centering
\caption{Pilot violation rates by sub-construct and three-level salience.}
\label{tab:pilot-subconstruct}
\begin{tabular}{lrr}
\toprule
Stratum & $n$ & Violation rate \\
\midrule
Authorization sub-construct & <<<AUTH_N>>>  & <<<AUTH_PCT>>>\% \\
State sub-construct         & <<<STATE_N>>> & <<<STATE_PCT>>>\% \\
Tool sub-construct          & <<<TOOL_N>>>  & <<<TOOL_PCT>>>\% \\
\midrule
High salience               & <<<HIGH_N>>>  & <<<HIGH_PCT>>>\% \\
Mid salience                & <<<MID_N>>>   & <<<MID_PCT>>>\% \\
Low salience                & <<<LOW_N>>>   & <<<LOW_PCT>>>\% \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The recall-ceiling caveat.}
Recall is at <<<RECALL_PCT>>>\% across the panel and items, with all but
<<<CNR_N>>> trials in the recalled cell. The HIUA construct in its strongest
form---action taken when the rule was not in mind at decision time---is
therefore \emph{unobserved} on this dataset. We interpret this as a property
of the items and panel jointly: the items use short, prominent prohibitions
that are paraphrase-stable under our recall probe, and the models in the
panel are sufficiently capable that none drops the rule from context within
the per-trial token budget. A construct version with harder recall probes
(asking for specific details---filename patterns, deadlines, scope
conditions---rather than the rule's existence) would likely populate this
cell, but at the cost of conflating measurement difficulty with construct
incidence. We return to this in the consequential-aspect discussion in
§\ref{sec:discussion}.

\paragraph{Comparison to baselines.}
The pilot's overall violation rate of <<<VIOLATION_PCT>>>\% sits in the
range reported by ToolEmu's outcome-only scoring (23.9\% safest-agent failure
rate~\cite{toolemu2024}) and Agent-SafetyBench (no model above 60\%
safety~\cite{agentsafetybench2024}), but our partition decomposes that rate
into the KBV cell that DriftBench measures~\cite{driftbench2026} and the
hallucination-type cells that HEAL probes via plan correctness rather than
prohibition crossing~\cite{heal2025}. An unpartitioned benchmark would
collapse the <<<N_CLEAN>>> trials into a single
<<<VIOLATION_PCT>>>\% aggregate, losing the within-trial mechanism the
$2 \times 4 \times 2$ partition recovers.

\paragraph{What the pilot establishes.}
First, the three-layer scoring stack is empirically separable, and the
recall scorer's substring layer alone would have missed <<<JUDGE_RESCUE_PCT>>>\%
of correct recalls (rescued by LLM-judge fallback)---direct evidence that a
benchmark using substring scoring alone would systematically inflate HIUA
estimates by misclassifying paraphrased recalls. Second, the partition
produces non-trivial occupancy in the violation-recalled cell on <<<N_MODELS_WITH_KBV>>>
of 13 models, operationally distinguishing it from compliance and from the
unobserved HIUA cell. Third, the salience contrast (<<<HIGH_PCT>>>\% high vs.\
<<<MID_PCT>>>\% mid vs.\ <<<LOW_PCT>>>\% low) is the panel's largest single
effect by an order of magnitude over the sub-construct contrast, with a
non-monotonic shape that the prior two-level design could not detect. The
pilot's specific HIUA-to-KBV ratio does not yet generalize, but the
partition is empirically realizable on the panel and items studied.
Limitations bound these claims and are discussed in §\ref{sec:discussion}.

% =========================================================================
% Replacement for §6 Discussion paragraphs that lean on pilot numbers
% =========================================================================

%% Replace the "What worked" paragraph in §6:

% \paragraph{What worked.}
% The three-layer scoring stack discriminated as designed: action-scorer regex
% and refusal-prefix layers handled <<<FAST_PATH_PCT>>>\% of trials, with
% LLM-judge fallback firing only on the ambiguous minority. The recall
% scorer's substring layer alone would have missed <<<JUDGE_RESCUE_PCT>>>\% of
% correct recalls (rescued by judge fallback), providing direct evidence that
% substring scoring alone would systematically inflate HIUA estimates by
% miscounting paraphrased recalls as recall failures. The
% $2 \times 4 \times 2$ partition produced non-trivial occupancy in the
% violation-recalled (KBV) cell while the violation-not-recalled (HIUA)
% cell remained empty, operationally locating the empirical violation mass on
% the lucid-violation side of the partition for this panel. The
% non-monotonic three-level salience contrast (<<<HIGH_PCT>>>\% vs.\ <<<MID_PCT>>>\%
% vs.\ <<<LOW_PCT>>>\%) was the panel's largest single effect, with the
% mid-salience pattern providing a finding the prior two-level design could
% not detect: rules buried as one of several mid-paragraph instructions are
% violated \emph{more} than rules tucked into tool-docstring corners.

%% Replace the "What didn't" paragraph in §6:

% \paragraph{What didn't.}
% Judge ensemble Cohen's $\kappa$ averaged <<<KAPPA_MEAN>>> across the three
% pairwise comparisons (full numbers in App.~\ref{app:pilot}); the
% substantive-validity argument is bounded by this level of agreement.
% Two of <<<N_TRIALS>>> trials failed at the agent layer
% (gpt-oss-20b emitted Python tool-call JSON on \texttt{tool-fileops} items
% even though no tools were declared, and the API rejected the wire format).
% These are arguably HIUA-relevant events---the model attempted a forbidden
% action---but were excluded from the headline rate because the wire-format
% failure prevented our scorer from inspecting the intended action.
% The single-seed design forecloses the variance-component decomposition that
% the §\ref{sec:design} reliability plan calls for: per-model rates lack
% standard errors and may not survive a full G-study. Recall remained at
% effective ceiling (<<<RECALL_PCT>>>\%), foreclosing direct observation of the
% violation-not-recalled (HIUA-strict) cell on this dataset.

% =========================================================================

\section{Pilot operationalization (extended)}
\label{app:pilot}

We expand here on the scoring-stack diagnostics and per-judge reliability
from §\ref{sec:pilot}.

\paragraph{Action scorer path distribution.}
Of the <<<N_CLEAN>>> clean trials, <<<REGEX_PCT>>>\% were scored by the regex
layer alone, <<<REFUSAL_PCT>>>\% by the refusal-prefix heuristic,
<<<JUDGE_RESOLVED_PCT>>>\% required the LLM-judge fallback for
both-patterns-match ambiguity, and <<<AMBIGUOUS_PCT>>>\% returned scorer
errors that the judge could not resolve. The fallback rate is low because
the action scorer's two cheap layers handle the majority of cases---the
design intent. The Harmony-format and curly-quote-aware extensions to the
refusal-prefix heuristic (added after observation of GPT-OSS family
responses) recovered <<<HARMONY_RECOVERED_N>>> trials that the original
parser had silently mis-classified.

\paragraph{Recall scorer path distribution.}
Of the <<<RECALLED_N>>> correctly-recalled trials, <<<SUBSTRING_PCT>>>\% were
caught by exact-substring match, while <<<JUDGE_YES_PCT>>>\% were rescued by
the LLM-judge fallback after the substring scorer had failed. This is the
operationalization basis for the substantive-validity claim in
§\ref{sec:pilot}: a benchmark using substring matching alone would
systematically misclassify <<<JUDGE_RESCUE_PCT>>>\% of correct recalls as
recall failures, which would inflate the HIUA estimate by counting
paraphrased-recall violations as not-recalled.

\paragraph{Judge ensemble reliability.}
The three-judge ensemble for the v2 pilot comprises Claude Sonnet 4.6,
Llama 3.3 70B Versatile (Groq), and Llama 4 Scout 17B (Groq). The pairwise
Cohen's $\kappa$ values are in Table~\ref{tab:judge-kappa-v2}.

\begin{table}[ht]
\centering
\caption{Pairwise judge agreement on hallucination-type labels (v2 pilot).}
\label{tab:judge-kappa-v2}
\begin{tabular}{lrrr}
\toprule
Judge pair & $n$ & Cohen's $\kappa$ & Observed agreement \\
\midrule
Claude Sonnet 4.6 $\times$ Llama 3.3 70B Versatile  & <<<PAIR1_N>>> & <<<PAIR1_KAPPA>>> & <<<PAIR1_OBS>>>\% \\
Claude Sonnet 4.6 $\times$ Llama 4 Scout 17B        & <<<PAIR2_N>>> & <<<PAIR2_KAPPA>>> & <<<PAIR2_OBS>>>\% \\
Llama 3.3 70B Versatile $\times$ Llama 4 Scout 17B  & <<<PAIR3_N>>> & <<<PAIR3_KAPPA>>> & <<<PAIR3_OBS>>>\% \\
\midrule
Mean of pairwise $\kappa$                            & --- & <<<KAPPA_MEAN>>> & --- \\
All-three-agree rate                                 & <<<N_CLEAN>>> & --- & <<<ALL_AGREE_PCT>>>\% \\
\bottomrule
\end{tabular}
\end{table}

The Claude Sonnet judge provides cross-family independence from the two Meta
Llama judges and operates without the daily-request-per-model cap that
constrained the v1 pilot's Llama 3.3 70B Versatile coverage. All
<<<N_CLEAN>>> trials carry valid labels from all three judges.

\paragraph{Substantive-validity caveat.}
The mean pairwise $\kappa$ of <<<KAPPA_MEAN>>> indicates
<<<KAPPA_QUALITATIVE>>> agreement; the substantive-validity argument is
correspondingly <<<SUBSTANTIVE_VERDICT>>>. The construct's empirical
residence---the cell partition of $\{$recall$\} \times \{$hallucination-type
or none$\} \times \{$violation$\}$---is realized in the data, but the rate
at which trials are assigned to the HIUA cell vs.\ the KBV cell is mediated
by the judge ensemble at the level of agreement reported above.

% =========================================================================
% Notes for compiler / fill-in pass
% =========================================================================

% After analyze.py --g-study completes, slot the following from its output:
%
%   - N_CLEAN, N_ERRORS = total clean trials, total errors
%   - HIGH_PCT, MID_PCT, LOW_PCT, HIGH_N, MID_N, LOW_N = salience table
%   - AUTH_PCT, STATE_PCT, TOOL_PCT, AUTH_N, STATE_N, TOOL_N = sub-construct table
#   - CR_N, CR_PCT, CNR_N, CNR_PCT, VR_N, VR_PCT, VNR_N, VNR_PCT = cell occupancy
%   - MAJ_AUTH, MAJ_STATE, MAJ_TOOL, MAJ_NONE = majority labels in VR cell
%   - HIUA_NUM, HIUA_PCT, KBV_NUM, KBV_PCT, HIUA_KBV_RATIO = construct estimates
%   - MODEL_TABLE_ROWS = generated from per-model table (see analyze.py output)
%   - REGEX_PCT, REFUSAL_PCT, JUDGE_RESOLVED_PCT, AMBIGUOUS_PCT = action scorer dist
%   - SUBSTRING_PCT, JUDGE_YES_PCT, JUDGE_RESCUE_PCT, RECALLED_N = recall scorer dist
%   - PAIR1_N, PAIR1_KAPPA, PAIR1_OBS, etc. = judge pair kappa table
%   - KAPPA_MEAN, ALL_AGREE_PCT = ensemble summary
%   - CROSS_PROVIDER_NOTE = prose summary of 4-model cross-provider replication
%   - VIOLATION_PCT, FAST_PATH_PCT, RECALL_PCT, HARMONY_RECOVERED_N = misc derived
%   - N_MODELS_WITH_KBV = count of models with ≥1 violation-recalled trial
%   - KAPPA_QUALITATIVE = "moderate" if 0.4-0.6, "substantial" if 0.6-0.8, etc.
%   - SUBSTANTIVE_VERDICT = "tightly bounded" / "moderately bounded" / etc.
